---
title: "00 — Load & describe human behavioral data"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    df-print: kable
    fig-width: 9
    fig-height: 5
execute:
  warning: false
  message: false
---

# Purpose

This notebook is the entry point for analysis of the **human** behavioral data
collected by the auto-psych outer loop on the `subjective_randomness` task. In
this task each trial shows two coin-flip sequences (H/T strings) and the
participant picks the one that looks "more random."

The data are organized as a set of **chains** (`run1`, `run2`, `run3`) — three
independent replications of the outer loop — each with three sequential
**rounds** (`experiment1`, `experiment2`, `experiment3`). Every (run, round)
cell is one freshly-recruited Prolific batch: 40 participants × 32 trials.

This notebook:

- merges all nine cells into one tidy trial-level table, giving every
  participant a **globally unique id** (the raw `participant_id` only runs 0–39
  *within* a cell, so ids collide across cells);
- reports basic diagnostics of **what participants saw** across runs and rounds:
  number of participants, collection-time, and stimulus lengths.

Choice biases and the stimulus-pair / sub-sequence reuse analyses live in
`01_response_analysis.qmd`, which reads the merged table this notebook saves.

```{r setup}
library(tidyverse)
library(here)
library(glue)
library(fs)

theme_set(theme_bw(base_size = 12))
options(dplyr.summarise.inform = FALSE)

# Root of the committed human-experiment results.
data_root <- here("data", "results", "human_experiment")
```

# Load & merge

We walk every `run*/subjective_randomness/experiment*/data/responses.csv`,
tagging each with its `run` and `round` (= experiment number), then build a
`participant_uid` that is unique across the whole corpus.

```{r load-responses}
response_files <- dir_ls(
  data_root,
  recurse = TRUE,
  glob = "*/data/responses.csv"
)

parse_cell <- function(path) {
  # path looks like .../runN/subjective_randomness/experimentM/data/responses.csv
  run   <- as.integer(str_match(path, "run(\\d+)")[, 2])
  round <- as.integer(str_match(path, "experiment(\\d+)")[, 2])

  read_csv(path, col_types = cols(
    participant_id = col_integer(),
    trial_index    = col_integer(),
    sequence_a     = col_character(),
    sequence_b     = col_character(),
    chose_left     = col_integer(),
    chose_right    = col_integer(),
    model          = col_character()
  )) |>
    mutate(run = run, round = round, .before = 1)
}

trials_raw <- map(response_files, parse_cell) |> list_rbind()

# `model` is blank for human responses; confirm and drop it.
stopifnot(all(is.na(trials_raw$model) | trials_raw$model == ""))

trials <- trials_raw |>
  select(-model) |>
  mutate(
    participant_uid = glue("r{run}_e{round}_p{participant_id}"),
    # which side / which sequence was chosen
    chosen_side = case_when(
      chose_left  == 1 & chose_right == 0 ~ "left",
      chose_right == 1 & chose_left  == 0 ~ "right",
      TRUE                                ~ NA_character_
    ),
    chosen_seq = if_else(chosen_side == "left", sequence_a, sequence_b),
    # sequence lengths
    len_a = str_length(sequence_a),
    len_b = str_length(sequence_b)
  ) |>
  relocate(participant_uid, .after = round)

glue("Merged {nrow(trials)} trials from {length(response_files)} cells ",
     "({n_distinct(trials$participant_uid)} unique participants).")
```

Sanity check: every (run, round) cell should have 40 participants × 32 trials =
1280 rows, and `participant_id` should be 0–39 within each cell.

```{r cell-shape}
trials |>
  group_by(run, round) |>
  summarise(
    n_participants = n_distinct(participant_id),
    n_trials       = n(),
    trials_per_pid = n() / n_distinct(participant_id),
    pid_min        = min(participant_id),
    pid_max        = max(participant_id),
    .groups = "drop"
  ) |>
  knitr::kable()
```

A few rows of the merged table:

```{r peek}
trials |>
  select(run, round, participant_uid, trial_index,
         sequence_a, sequence_b, chosen_side, chosen_seq) |>
  slice_head(n = 6) |>
  knitr::kable()
```

# Diagnostics: what participants saw

## Number of participants

```{r n-participants}
trials |>
  distinct(run, round, participant_id) |>
  count(run, round, name = "n_participants") |>
  pivot_wider(names_from = round, values_from = n_participants,
              names_prefix = "round ") |>
  knitr::kable(caption = "Participants per chain (run) × round")
```

```{r n-participants-total}
glue("Total unique participants across all chains and rounds: ",
     "{n_distinct(trials$participant_uid)} ",
     "(= {n_distinct(trials$run)} runs × {n_distinct(trials$round)} rounds × 40).")
```

## Collection time

**Caveat.** The committed data do **not** include per-participant response
times or task durations — `responses.csv` has no timing columns. The only
timing signal is each cell's `data/observability.log`, which polls Prolific
every ~30 s and records the cumulative number of completed submissions. From
that we can recover, per cell, **how long the batch took to fill** and the
**rate at which submissions arrived** — but not how long any individual
participant spent on the task.

```{r parse-logs}
log_files <- dir_ls(data_root, recurse = TRUE, glob = "*/data/observability.log")

parse_log <- function(path) {
  run   <- as.integer(str_match(path, "run(\\d+)")[, 2])
  round <- as.integer(str_match(path, "experiment(\\d+)")[, 2])

  lines <- read_lines(path)
  poll  <- str_match(lines,
    "^\\[(\\d{4}-\\d{2}-\\d{2}T\\d{2}:\\d{2}:\\d{2}Z)\\].*completed=(\\d+) target=(\\d+)")
  poll  <- poll[!is.na(poll[, 1]), , drop = FALSE]
  if (nrow(poll) == 0) return(NULL)

  tibble(
    run       = run,
    round     = round,
    timestamp = ymd_hms(poll[, 2]),
    completed = as.integer(poll[, 3]),
    target    = as.integer(poll[, 4])
  )
}

polls <- map(log_files, parse_log) |> list_rbind() |>
  group_by(run, round) |>
  mutate(elapsed_min = as.numeric(timestamp - min(timestamp), units = "mins")) |>
  ungroup()
```

Per-cell collection duration (first poll → batch full):

```{r collection-duration}
collection_summary <- polls |>
  group_by(run, round) |>
  summarise(
    target          = first(target),
    final_completed = max(completed),
    duration_min    = max(elapsed_min[completed == max(completed)]),
    .groups = "drop"
  )

collection_summary |>
  knitr::kable(
    digits = 1,
    col.names = c("run", "round", "target", "completed", "duration (min)"),
    caption = "Time to fill each 40-participant batch"
  )
```

```{r collection-curves}
polls |>
  mutate(cell = glue("run {run} · round {round}")) |>
  ggplot(aes(x = elapsed_min, y = completed, color = factor(round))) +
  geom_step() +
  facet_wrap(~ run, labeller = label_both) +
  scale_color_brewer(palette = "Dark2", name = "round") +
  labs(
    x = "Minutes since collection start",
    y = "Cumulative completed submissions",
    title = "Submission accrual per chain × round",
    subtitle = "From Prolific poll logs — batch fill rate, not individual task time"
  )
```

## Stimulus lengths

Each stimulus is a string of H/T characters. Below is the distribution of
sequence lengths (pooling `sequence_a` and `sequence_b`), shown per round since
the design generator can change stimulus characteristics across rounds.

```{r stim-lengths}
seq_lengths <- trials |>
  distinct(run, round, sequence_a, sequence_b) |>
  pivot_longer(c(sequence_a, sequence_b),
               names_to = "slot", values_to = "sequence") |>
  distinct(run, round, sequence) |>
  mutate(len = str_length(sequence))

seq_lengths |>
  ggplot(aes(x = len, fill = factor(round))) +
  geom_histogram(binwidth = 1, color = "white") +
  facet_grid(round ~ run, labeller = label_both) +
  scale_fill_brewer(palette = "Dark2", guide = "none") +
  scale_x_continuous(breaks = scales::breaks_width(2)) +
  labs(x = "Sequence length (characters)", y = "Number of distinct sequences",
       title = "Stimulus length distribution by chain × round")
```

```{r stim-lengths-table}
seq_lengths |>
  group_by(run, round) |>
  summarise(
    n_distinct_seq = n(),
    len_min  = min(len),
    len_median = median(len),
    len_max  = max(len),
    .groups = "drop"
  ) |>
  knitr::kable(digits = 1, caption = "Distinct sequences and length range per cell")
```

# Save

We persist the merged trial-level table for downstream analysis notebooks.

```{r save}
out_dir <- here("analysis", "behavioral", "data")
dir_create(out_dir)

write_rds(trials, path(out_dir, "human_trials.rds"))
write_csv(trials, path(out_dir, "human_trials.csv"))

glue("Saved {nrow(trials)} trials ({n_distinct(trials$participant_uid)} participants) ",
     "to analysis/behavioral/data/human_trials.{{rds,csv}}.")
```

# Notes

- **Unique participant ids.** `participant_uid = r{run}_e{round}_p{participant_id}`.
  Each (run, round) cell is an independent Prolific batch, so the raw 0–39
  `participant_id` is only unique within a cell. Do **not** merge cells on
  `participant_id` alone.
- **Use `data/responses.csv`, not `model_loop/responses.csv`.** The per-round
  `data/responses.csv` is the fresh batch for that round. The sibling
  `model_loop/responses.csv` accumulates earlier rounds' data *under colliding
  participant ids* (and adds model feature columns); it is for model fitting,
  not participant-level description.
- **No per-participant timing in the committed data.** Response times / task
  durations are not in `responses.csv`. Collection-time diagnostics here come
  from the Prolific poll logs and describe batch fill rate only.
