---
title: "01 — Response analysis: choice biases & stimulus reuse"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    df-print: kable
    fig-width: 9
    fig-height: 5
execute:
  warning: false
  message: false
---

# Purpose

Building on `00_load_human_data.qmd`, this notebook analyses the **choices**
participants made on the `subjective_randomness` task ("Which sequence looks
more random?") and characterises the **stimulus structure** of the corpus.

Two parts:

1. **Choice biases.** Simple "proportion choosing the side that follows a rule"
   summaries for four candidate biases — side, number-of-heads, length, and
   alternation — across runs and rounds.
2. **Stimulus reuse**, with H/T relabelling consolidated (see the note below):
   pair-level reuse and single-sequence ("side") reuse across the corpus.

```{r setup}
library(tidyverse)
library(here)
library(glue)

theme_set(theme_bw(base_size = 12))
options(dplyr.summarise.inform = FALSE)

trials <- read_rds(here("analysis", "behavioral", "data", "human_trials.rds"))
```

## A note on H/T symmetry

Heads and Tails are arbitrary labels, so for the **stimulus-structure** analyses
we treat a sequence as equivalent to its H↔T relabelling (e.g. `HT` ≡ `TH`,
`HHT` ≡ `TTH`). We canonicalise each sequence to the lexicographically smaller
of itself and its complement:

```{r ht-helpers}
complement   <- function(s) chartr("HT", "TH", s)
canon_seq    <- function(s) pmin(s, complement(s))   # H/T-canonical single sequence

# number of H↔T alternations (switches between adjacent symbols)
n_alternations <- function(s) {
  ch <- str_split(s, "")
  map_int(ch, \(x) if (length(x) < 2) 0L else sum(x[-1] != x[-length(x)]))
}
```

We apply this **per sequence** for the single-sequence analysis, and to the
**whole pair** (flipping H/T on both sides together) for the pair analysis —
because "which looks more random" is invariant to relabelling the entire
display, but *not* to flipping just one side. We do **not** apply reversal
symmetry (reading a sequence backwards); ask if that's wanted too.

# Choice biases

For each trial we compute features of both sequences, then ask, for each
candidate bias, **what proportion of choices followed the rule** — choosing the
side with the higher value of that feature. Trials where the two sequences tie
on a feature carry no information about that bias and are dropped from its
denominator. We summarise per participant (the unit of analysis), then average
across participants within each round.

```{r trial-features}
# Per-sequence features, joined back for side a and side b.
seq_feat <- trials |>
  distinct(sequence_a, sequence_b) |>
  pivot_longer(everything(), values_to = "seq") |>
  distinct(seq) |>
  mutate(
    len     = str_length(seq),
    n_heads = str_count(seq, "H"),
    n_alt   = n_alternations(seq),
    alt_rate = n_alt / (len - 1)            # len >= 2, so denominator >= 1
  )

bias_trials <- trials |>
  select(-len_a, -len_b) |>   # recomputed (with more features) from seq_feat below
  left_join(seq_feat |> rename_with(~ paste0(.x, "_a")), by = c("sequence_a" = "seq_a")) |>
  left_join(seq_feat |> rename_with(~ paste0(.x, "_b")), by = c("sequence_b" = "seq_b")) |>
  mutate(
    chose_a = chosen_side == "left",   # sequence_a is shown on the left
    # For each rule: did the chosen side have the strictly higher feature value?
    # NA when the two sides tie on that feature (excluded from that bias).
    follow_side  = chose_left == 1,                       # "rule" = chose left
    follow_heads = case_when(n_heads_a == n_heads_b ~ NA,
                             TRUE ~ (n_heads_a > n_heads_b) == chose_a),
    follow_len   = case_when(len_a == len_b ~ NA,
                             TRUE ~ (len_a > len_b) == chose_a),
    follow_alt   = case_when(alt_rate_a == alt_rate_b ~ NA,
                             TRUE ~ (alt_rate_a > alt_rate_b) == chose_a)
  )
```

```{r bias-per-participant}
bias_long <- bias_trials |>
  select(run, round, participant_uid,
         side = follow_side, n_heads = follow_heads,
         length = follow_len, alternation = follow_alt) |>
  pivot_longer(c(side, n_heads, length, alternation),
               names_to = "bias", values_to = "follow") |>
  mutate(bias = factor(bias, levels = c("side", "n_heads", "length", "alternation")))

# Per-participant proportion following each rule (ties excluded via na.rm).
# Each participant belongs to exactly one (run, round) cell, so this is one
# value per participant per bias.
bias_pp <- bias_long |>
  group_by(run, round, participant_uid, bias) |>
  summarise(prop = mean(follow, na.rm = TRUE),
            n_used = sum(!is.na(follow)), .groups = "drop")
```

## Topline summary

Pooled across all three chains and three rounds, with the **participant as the
unit of analysis**: the per-participant proportion of rule-following choices,
averaged over participants, with a t-based 95% CI. `0.5` = no bias.

```{r bias-topline}
bias_topline <- bias_pp |>
  filter(is.finite(prop)) |>
  group_by(bias) |>
  summarise(
    n_participants = n(),
    estimate = mean(prop),
    se       = sd(prop) / sqrt(n()),
    ci_lo    = estimate - qt(0.975, n() - 1) * se,
    ci_hi    = estimate + qt(0.975, n() - 1) * se,
    .groups = "drop"
  )

bias_topline |>
  transmute(
    bias,
    n_participants,
    `proportion [95% CI]` = sprintf("%.3f [%.3f, %.3f]", estimate, ci_lo, ci_hi)
  ) |>
  knitr::kable(caption = paste(
    "Topline choice biases, pooled across all runs and rounds",
    "(participant-level mean ± 95% CI; 0.5 = no bias)."
  ))
```

```{r bias-topline-text}
pct <- function(b) {
  r <- bias_topline |> filter(bias == b)
  sprintf("%.1f%%, 95%% CI [%.1f%%, %.1f%%]", 100 * r$estimate, 100 * r$ci_lo, 100 * r$ci_hi)
}
glue(
  "Across all {n_distinct(bias_pp$participant_uid)} participants, choices showed ",
  "no side bias ({pct('side')} chose the left sequence), ",
  "a negligible heads bias ({pct('n_heads')} chose the sequence with more heads), ",
  "a modest length bias ({pct('length')} chose the longer sequence), ",
  "and a clear alternation bias ({pct('alternation')} chose the higher-alternation sequence)."
)
```

Each rule is scored so that **0.5 = no bias**; the dashed line marks chance.

- **side** — proportion choosing the *left* sequence.
- **n_heads** — proportion choosing the sequence with *more heads*.
- **length** — proportion choosing the *longer* sequence.
- **alternation** — proportion choosing the sequence with the higher
  *alternation rate* (switches per adjacent pair). The classic
  representativeness prediction is a strong alternation bias (> 0.5).

```{r bias-summary-plot}
bias_summary <- bias_pp |>
  group_by(round, bias) |>
  summarise(
    mean_prop = mean(prop),
    se        = sd(prop) / sqrt(n()),
    n_part    = n(),
    .groups = "drop"
  )

bias_summary |>
  ggplot(aes(x = factor(round), y = mean_prop, color = factor(round))) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "grey50") +
  geom_pointrange(aes(ymin = mean_prop - 1.96 * se,
                      ymax = mean_prop + 1.96 * se)) +
  facet_wrap(~ bias, nrow = 1) +
  scale_color_brewer(palette = "Dark2", name = "round") +
  scale_y_continuous(limits = c(0, 1)) +
  labs(x = "round", y = "Proportion of choices following rule",
       title = "Choice biases across rounds (mean ± 95% CI over participants)",
       subtitle = "0.5 = no bias; pooled over the three chains")
```

Per (run, round) table of the same proportions:

```{r bias-table}
bias_pp |>
  group_by(run, round, bias) |>
  summarise(prop = mean(prop), .groups = "drop") |>
  pivot_wider(names_from = bias, values_from = prop) |>
  knitr::kable(digits = 2,
               caption = "Mean proportion of choices following each rule, per cell")
```

# Stimulus pair reuse

Each round presents **32 distinct (unordered) pairs**, every participant seeing
each once. Here we consolidate pairs under whole-display H/T relabelling and ask
how often the same (consolidated) pair recurs across the nine cells.

```{r pair-canon}
pair_canon <- trials |>
  distinct(run, round, sequence_a, sequence_b) |>
  mutate(
    # sorted within-pair (order-independent), and its whole-pair H/T complement
    lo = pmin(sequence_a, sequence_b),
    hi = pmax(sequence_a, sequence_b),
    raw_pair  = paste(lo, hi, sep = " | "),
    comp_lo   = pmin(complement(sequence_a), complement(sequence_b)),
    comp_hi   = pmax(complement(sequence_a), complement(sequence_b)),
    comp_pair = paste(comp_lo, comp_hi, sep = " | "),
    pair_canon = pmin(raw_pair, comp_pair)
  ) |>
  select(run, round, pair_canon)
```

```{r pair-consolidation-note}
n_raw <- trials |>
  distinct(sequence_a, sequence_b) |>
  mutate(k = paste(pmin(sequence_a, sequence_b), pmax(sequence_a, sequence_b))) |>
  pull(k) |> n_distinct()
n_canon <- n_distinct(pair_canon$pair_canon)

glue("Distinct pairs: {n_raw} ignoring H/T symmetry -> {n_canon} after ",
     "consolidating each pair with its H/T complement.")
```

How many of the nine cells does each consolidated pair appear in?

```{r pair-reuse-plot}
pair_reuse <- pair_canon |>
  distinct(run, round, pair_canon) |>
  count(pair_canon, name = "n_cells")

pair_reuse |>
  count(n_cells, name = "n_pairs") |>
  ggplot(aes(x = factor(n_cells), y = n_pairs)) +
  geom_col(fill = "steelblue") +
  geom_text(aes(label = n_pairs), vjust = -0.4, size = 3.5) +
  labs(x = "Number of (run × round) cells the pair appears in",
       y = "Number of distinct stimulus pairs",
       title = "Stimulus-pair reuse across the corpus (H/T-consolidated)",
       subtitle = glue("{nrow(pair_reuse)} distinct consolidated pairs total")) +
  expand_limits(y = max(count(pair_reuse, n_cells)$n) * 1.1)
```

Pairs that recur in more than one cell — candidates for comparing responses to
the same stimulus across conditions:

```{r pair-reuse-table}
pair_reuse |>
  filter(n_cells > 1) |>
  arrange(desc(n_cells), pair_canon) |>
  knitr::kable(col.names = c("pair (H/T-canonical)", "n cells"),
               caption = "Consolidated stimulus pairs appearing in >1 cell")
```

# Single-sequence (side) reuse

A pair has two sides, each a full sequence. Here we ignore pairing and ask how
often each individual (H/T-canonical) sequence appears as a stimulus, and in how
many cells — i.e. which *sequences* the design reused, regardless of partner.

```{r side-reuse}
side_seqs <- trials |>
  distinct(run, round, sequence_a, sequence_b) |>
  pivot_longer(c(sequence_a, sequence_b), values_to = "seq") |>
  mutate(seq_canon = canon_seq(seq)) |>
  distinct(run, round, seq_canon)

seq_reuse <- side_seqs |>
  count(seq_canon, name = "n_cells") |>
  mutate(len = str_length(seq_canon))

glue("{n_distinct(side_seqs$seq_canon)} distinct H/T-canonical sequences used ",
     "as sides across the corpus.")
```

```{r side-reuse-plot}
seq_reuse |>
  count(n_cells, name = "n_seqs") |>
  ggplot(aes(x = factor(n_cells), y = n_seqs)) +
  geom_col(fill = "darkorange") +
  geom_text(aes(label = n_seqs), vjust = -0.4, size = 3.5) +
  labs(x = "Number of (run × round) cells the sequence appears in (as either side)",
       y = "Number of distinct sequences",
       title = "Single-sequence reuse across the corpus (H/T-consolidated)") +
  expand_limits(y = max(count(seq_reuse, n_cells)$n) * 1.1)
```

The most-reused individual sequences:

```{r side-reuse-table}
seq_reuse |>
  arrange(desc(n_cells), len, seq_canon) |>
  slice_head(n = 20) |>
  knitr::kable(col.names = c("sequence (H/T-canonical)", "n cells", "length"),
               caption = "Most-reused individual sequences (by cells appearing in)")
```

# Notes

- **Bias scoring.** Each bias is "proportion of non-tie trials whose chosen side
  has the higher feature value," scored so 0.5 = no bias. `side` is the special
  case of proportion-choosing-left. Participant is the unit of analysis.
- **H/T consolidation** is applied per-sequence for the side analysis and
  per-whole-pair for the pair analysis; reversal symmetry is not applied.
