---
title: "02 — Best-fitting theories: mega-analytic model comparison"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    df-print: kable
    fig-width: 9
    fig-height: 5
execute:
  warning: false
  message: false
---

# Purpose

This is the **mega-analytic model comparison** for the paper. We pool the human
behavioral data (3 runs × 3 rounds × 40 participants × 32 trials = 11,520
forced-choice trials) and ask which theory best accounts for the collective
dataset.

The model set (`mega_models.json`):

- **4 canonical seed theories** the outer loop started from — *prototype
  similarity*, *encoding compressibility*, *Bayesian diagnosticity*, *window
  typicality*.
- **4 overall run winners** the loop discovered — the best model of each run's
  final round (run 2 ended in a near-tie, so it contributes two).

Each is a population-level PyMC model fit with the project's own inference
harness (`src/models/pymc_inference.py`). The fitting (`fit_mega_models.py`,
run on Sherlock) produces two schemes:

- **`mega`** — every model fit on **all** the data, scored in-sample
  (ELPD-LOO + RMSE/R²). The headline "collective fit."
- **`heldout`** — selection-debiased: each winner was *discovered* on one run, so
  we refit it (and the seeds) only on the **other two runs**. We report the fit
  on those two training runs (`heldout_train`, ELPD-LOO) and genuine
  out-of-sample prediction of the held-out run (`heldout_oos`).

```{r setup}
library(tidyverse)
library(here)
library(glue)

theme_set(theme_bw(base_size = 12))
options(dplyr.summarise.inform = FALSE)

metrics_path <- here("analysis", "behavioral", "data", "mega_model_metrics.csv")
preds_path   <- here("analysis", "behavioral", "data", "mega_model_predictions.csv")
```

```{r guard}
#| echo: false
#| results: asis
if (!file.exists(metrics_path) || !file.exists(preds_path)) {
  cat(
    "::: callout-important\n",
    "## Fit outputs not found\n\n",
    "Run the fitting step first — it writes `mega_model_metrics.csv` and ",
    "`mega_model_predictions.csv` to `analysis/behavioral/data/`:\n\n",
    "```bash\n",
    "sbatch analysis/behavioral/slurm/run_mega_comparison.sbatch   # on Sherlock\n",
    "python analysis/behavioral/fit_mega_models.py                 # or locally\n",
    "```\n",
    ":::\n",
    sep = ""
  )
  knitr::knit_exit()
}
```

```{r load}
metrics <- read_csv(metrics_path, show_col_types = FALSE) |>
  mutate(kind = factor(kind, levels = c("seed", "winner")))
preds <- read_csv(preds_path, show_col_types = FALSE)

# Headline comparison = the mega scheme (all data, in-sample).
mega <- metrics |> filter(scheme == "mega", eval == "in_sample")
model_levels <- mega |> arrange(elpd_rank) |> pull(label)

preds_mega <- preds |> filter(scheme == "mega", eval == "in_sample")
```

# Mega comparison (all data)

Models ordered by ELPD-LOO (higher = better). `Δelpd` is the gap from the best
model; `dse` its standard error.

```{r table}
mega |>
  transmute(
    Model = label, Kind = kind, Params = n_params,
    `ELPD-LOO` = round(elpd_loo, 1), `± SE` = round(elpd_se, 1),
    `Δelpd` = round(elpd_diff, 1), dse = round(elpd_dse, 1),
    RMSE = round(rmse, 3), `R² trial` = round(r2_trial, 3),
    `R² stim` = round(r2_stimulus, 3), `Acc.` = round(accuracy, 3),
    `LOO ok` = !loo_unreliable
  ) |>
  knitr::kable(caption = "Mega-analytic model comparison on the pooled human data")
```

```{r elpd-plot}
mega |>
  mutate(label = factor(label, levels = rev(model_levels))) |>
  ggplot(aes(x = elpd_loo, y = label, color = kind)) +
  geom_pointrange(aes(xmin = elpd_loo - elpd_se, xmax = elpd_loo + elpd_se)) +
  scale_color_manual(values = c(seed = "grey50", winner = "firebrick"), name = NULL) +
  labs(x = "ELPD-LOO (higher is better)", y = NULL,
       title = "Predictive fit of each theory on the pooled data") +
  theme(legend.position = "bottom")
```

Difference from the best model (± 2 dse). A model whose interval clears 0 is
reliably worse than the best.

```{r delpd-plot}
mega |>
  filter(elpd_rank > 0) |>
  mutate(label = factor(label, levels = rev(model_levels))) |>
  ggplot(aes(x = elpd_diff, y = label, color = kind)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  geom_pointrange(aes(xmin = elpd_diff - 2 * elpd_dse, xmax = elpd_diff + 2 * elpd_dse)) +
  scale_color_manual(values = c(seed = "grey50", winner = "firebrick"), name = NULL) +
  labs(x = "Δ ELPD-LOO from best model (± 2 dse)", y = NULL,
       title = "Gap to the best-fitting theory") +
  theme(legend.position = "bottom")
```

```{r rmse-r2}
mega |>
  select(label, kind, rmse, r2_trial, r2_stimulus) |>
  pivot_longer(c(rmse, r2_trial, r2_stimulus), names_to = "metric", values_to = "value") |>
  mutate(metric = recode(metric, rmse = "RMSE (lower better)",
                         r2_trial = "R² (trial-level)", r2_stimulus = "R² (stimulus-level)"),
         label = factor(label, levels = rev(model_levels))) |>
  ggplot(aes(x = value, y = label, fill = kind)) +
  geom_col() +
  facet_wrap(~ metric, scales = "free_x") +
  scale_fill_manual(values = c(seed = "grey60", winner = "firebrick"), name = NULL) +
  labs(x = NULL, y = NULL, title = "Point-prediction fit metrics") +
  theme(legend.position = "bottom")
```

## Stimulus-level calibration

Empirical vs. predicted P(choose left) for each distinct ordered stimulus, for
the best model of each kind (best seed, best winner). On-diagonal = calibrated.

```{r calibration}
best_by_kind <- mega |> group_by(kind) |> slice_min(elpd_rank, n = 1) |>
  ungroup() |> select(name, label, kind)

preds_mega |>
  semi_join(best_by_kind, by = c("model" = "name")) |>
  group_by(model, sequence_a, sequence_b) |>
  summarise(obs = mean(chose_left), pred = mean(p_left), .groups = "drop") |>
  left_join(best_by_kind, by = c("model" = "name")) |>
  ggplot(aes(x = pred, y = obs)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey50") +
  geom_point(aes(color = kind), alpha = 0.4, size = 1) +
  facet_wrap(~ label) +
  scale_color_manual(values = c(seed = "grey40", winner = "firebrick"), guide = "none") +
  coord_equal(xlim = c(0, 1), ylim = c(0, 1)) +
  labs(x = "Predicted P(choose left)", y = "Empirical proportion choosing left",
       title = "Stimulus-level calibration (best seed vs. best winner)")
```

# Held-out generalization (selection-debiased)

The winners were *discovered* on the human data, so their mega-scheme edge over
the seeds is partly selection optimism. Here we remove it: for each winner we
refit it — and the seeds — **only on the two runs it did not come from**, and
compare on identical data.

## Fit on the two non-home runs (`heldout_train`)

This is the debiased comparison: within each held-out run, do the winner(s)
discovered elsewhere still out-fit the seeds on data that played no role in
selecting them? ELPD-LOO on the two training runs.

```{r heldout-train}
ho_train <- metrics |> filter(scheme == "heldout", eval == "heldout_train")

ho_train |>
  group_by(held_out_run) |>
  mutate(rank_in_group = rank(elpd_rank)) |>
  ungroup() |>
  mutate(facet = glue("held out run {held_out_run}\n(fit on the other two)")) |>
  ggplot(aes(x = elpd_diff, y = reorder(label, -elpd_diff), color = kind)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  geom_pointrange(aes(xmin = elpd_diff - 2 * elpd_dse, xmax = elpd_diff + 2 * elpd_dse)) +
  facet_wrap(~ facet, scales = "free_y") +
  scale_color_manual(values = c(seed = "grey50", winner = "firebrick"), name = NULL) +
  labs(x = "Δ ELPD-LOO from best in group (± 2 dse)", y = NULL,
       title = "Debiased fit: winners vs seeds on the runs that didn't produce them") +
  theme(legend.position = "bottom")
```

```{r heldout-train-table}
ho_train |>
  filter(kind == "winner") |>
  transmute(`Held-out run` = held_out_run, Winner = label,
            `ELPD rank (of group)` = elpd_rank,
            `Δelpd vs best` = round(elpd_diff, 1), dse = round(elpd_dse, 1),
            `R² stim` = round(r2_stimulus, 3)) |>
  knitr::kable(caption = "Each winner's standing among the seeds when refit on its two non-home runs (rank 0 = best)")
```

## Out-of-sample prediction of the held-out run (`heldout_oos`)

Genuine out-of-sample: models fit on two runs, predicting the third. (For a
winner's *own* run this still reflects that run's role in its selection, so read
the seeds' OOS curve as the reference and the winners as "does the discovered
form transfer.") Stimulus-level R² on the held-out run.

```{r heldout-oos}
metrics |>
  filter(scheme == "heldout", eval == "heldout_oos") |>
  mutate(facet = glue("predict run {held_out_run}")) |>
  ggplot(aes(x = r2_stimulus, y = reorder(label, r2_stimulus), fill = kind)) +
  geom_col() +
  facet_wrap(~ facet, scales = "free_y") +
  scale_fill_manual(values = c(seed = "grey60", winner = "firebrick"), name = NULL) +
  labs(x = "Out-of-sample R² (stimulus-level) on the held-out run", y = NULL,
       title = "Generalization to a run excluded from fitting") +
  theme(legend.position = "bottom")
```

# Notes & caveats

- **ELPD-LOO** is PSIS-LOO (`arviz.loo`) on each model's pointwise
  log-likelihood; `dse` is the SE of the pairwise difference (`arviz.compare`).
  `LOO ok` flags unreliable PSIS (high Pareto-k).
- **RMSE / Brier** and **R²** use posterior-mean `p_left` vs. observed 0/1
  choice. Stimulus-level R² (predicted vs. empirical choice proportion per
  distinct ordered stimulus) is the most interpretable "does it predict the
  choice curve" number.
- **Why a held-out scheme.** The mega scheme is in-sample for the winners (they
  were discovered on these data); ELPD-LOO controls in-fit overfitting but not
  model-*selection* overfitting. `heldout_train` is the clean debiased test —
  each winner refit on the two runs that had no role in selecting it, compared to
  seeds on identical data.
- **Pooling.** One population-level fit per model (the models have no
  participant-level parameters). Run/round enter only the held-out splits and the
  diagnostics, not the within-fit likelihood.
